---
title: "Basic Processing Workflow"
author: "Mark Felice"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basic Processing Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
## Summary
This vignettes shows an example workflow for basic importing and processing of PLFA data from raw named peak lists to lipid and biomarker concentrations and mol %.

## Processing
Import packages
```{r, echo = T, results = 'hide', message = F}
library(plfaR)  # PLFA processing
library(tidyverse)  # Basic data wrangling
library(readxl)  # Import Excel files
```

Import sample metadata. At minimum, you should have the following columns:
* Batch
* DataFileName
* SampleType
* SampleWt
You can also include any additional data you will want for analysis.
```{r}
stm_1415 <- read_xls(path = paste0('C:\\Users\\Mark\\Dropbox\\',                                                                          'umn_gutknecht_postdoc\\',
                                   'spruce_project\\', 
                                   'plfa_13c\\data\\',
                                   'SPRUCE DATA Nov 2016_all the ',
                                   'final calculations.xls'), 
                     sheet = 'Stats Ready Data Values', na = 'NA',
                     range = cell_cols('A:I')) %>%
  rename(BatchDataFileName = `Sample ID`) %>%
  mutate(Month = factor(Month, levels = c('June', 'July', 'August', 
                                          'Sept', 'October'))) %>%
  select(c(BatchDataFileName, Plot, DepthInterval = `Depth Increment`, 
           CO2EventualTreatment = Treatment, Temp = `Soil Probe Temp`, 
           GWCPerc = `% Gravimetric Water Content`)) # Excluded Year and Month to avoid dups in merge

c_types1 <- c('text', 'text', 'text', 'text', 'numeric', 'numeric', 'text', 
              'numeric', 'text', 'text', 'numeric', 'numeric', 'numeric',
              'text', 'numeric')
md_1415 <- read_xls(path = paste0('C:\\Users\\Mark\\Dropbox\\',
                                  'umn_gutknecht_postdoc\\',
                                  'spruce_project\\', 
                                  'plfa_13c\\data\\',
                                  'SPRUCE DATA Nov 2016_all the ',
                                  'final calculations.xls'), 
                    sheet = 'Data for pivot tables', na = 'NA',
                    col_types = c_types1) %>%
  mutate(TopCoreDepthcm = as.numeric(if_else(str_extract(`Depth Increment (cm)`, 
                                                         '(?<=\\()[0-9]+') == '1', 
                                             '0', 
                                             paste0('-', str_extract(`Depth Increment (cm)`, 
                                                                     '(?<=\\()[0-9]+')
                                             )
  )
  ),
  BottomCoreDepthcm = as.numeric(paste0('-', 
                                        str_extract(`Depth Increment (cm)`, 
                                                    '(?<=, )[0-9]+'))),
  MidCoreDepthcm = TopCoreDepthcm + (BottomCoreDepthcm - TopCoreDepthcm)/2) %>%
  rename(BatchDataFileName = `ID_with Batch ID`) %>%
  distinct(BatchDataFileName, .keep_all = TRUE) %>% # cut metadata repeated for each biomarker
  left_join(stm_1415, by = c('BatchDataFileName')) %>%
  select(c(DataFileName = Name, Batch = `Batch ID`, SampleType = Category, 
           Month, Year, Plot, DepthInterval, MidCoreDepthcm, 
           SampleWt = `Weight (g)`, Temp, GWCPerc))
# Cell M2261 is blank, gives warning about expecting numeric

# Clean up workspace
rm(c_types1, stm_1415)
```

Import your PLFA data. This should be your named peak list, and should be an Excel file with the following formatting:
* Either single tab OR peak list data on a worksheet named 'named_peaks'
* File should have the following columns with headers:
  + Batch
  + DataFileName
  + RetTimeSecs
  + MajorHeightnA
  + TotalPeakArea1
  + DisplayDelta1
  + Name

```{r}
# Define file path
source_dir <- 'C:\\Users\\Mark\\Desktop\\temp plfa practice - can delete whenever'  
source_file <- 'spruce_2015_b14_example.xlsx'

# Imported data is dataframe with cols listed above
peak_list <- import_batch_base(paste0(source_dir, '/', source_file))
```

Run a QC tests on the data. If there are any issues with the data, an error message will be displayed indicating the issue. This also generates a list containing summary dataframes showing 1) samples with duplicate lipids (and which lipids), 2) samples missing standard peaks (13:0, 16:0*, or 19:0), and 3) the frequency (fraction) with which each lipid is present in a sample.

*NOTE: Although 16:0 isn't technically a standard, it is present in Sphagnum, so should be present in any peat samples. This can likely be ignored if working with mineral soils.
```{r}
qc_df <- quality_check_base(peak_list)
```

```{r, echo = T, results = 'hide', message = F}
# QC summary dataframes can be accessed individually
qc_df[[1]][['duplicate_lipids']]
qc_df[[1]][['missing_lipids']]
qc_df[[1]][['lipid_frequency']]
```

To process the data, we first convert peak areas to nmol/g soil
* In addition to supplying the peak list, we need to supply a vector containing the sample names for the blanks (should only have 13:0 and 19:0) and standards (ie. the samples on which you want to base kval calculations.

NOTE: process_peak_area_base() also has several other parameters that can be changed, but which are constant in our lab.
```{r}
# Create vector of filenames of standards (can also manually supply)
stds <- c('13 Standard 1.raw', '4x concentraed 19 standard.raw')

# Blanks represent peak area subtracted for 13:0 and 19:0, so input a named 
# list with a vector for each lipid to subtract
blanks <- list( '13:0' = stds,
                '19:0' = c('Control 1 for plot 16 17 and 7.raw',
                           'Control 1 for plots 19 20  21.raw',
                           'Control 2 for plot 16 17 and 7.raw',
                           'Control 2 for plots 19 20  21.raw)'))

# Convert the peak areas to concentration (nmol/g dry soil)
conc_df <- process_peak_area_base(peak_list, 
                                  blanks = blanks, 
                                  standard_fnames = stds,
                                  soil_wt_df = md_1415)
```

Next, we can calculate the concentrations of groups of indicator lipids as well as their mol %.
```{r}
# I'm removing 16:0 and 19:0 peaks for this analysis
# use the output from concentration calcs as the input
indicators_df <- filter(conc_df, !Name %in% c('16:0', '19:0')) %>%
                 calculate_indicators_base(soil_wt_df = md_1415)

```

These dataframes can now be analyzed with your preferred stats pipeline.

## Apendix
The following are the groupings of lipids used for each biomarker group.
```{r, echo = F}
indicator_list <- list(f_lipids = c('18:1 w9c', '18:2 w6,9c'),
                       b_lipids = c('15:0 iso', '15:0 anteiso',
                                    '16:1 w7c', '16:0 10me',
                                    '18:1 w9c', '18:1 w9t',
                                    '18:2 w6,9c', '18:0 10me', '19:0 cyclo'),
                       gram_pos_lipids = c('15:0 iso', '15:0 anteiso'),
                       gram_neg_lipids = c('16:1 w7c', '18:1 w9t'),
                       actino_lipids = c('16:0 10me', '18:0 10me'),
                       anaerobe_lipids = c('19:0 cyclo'),
                       protozoa = c('20:4 w6,9,12,15'),
                       total_biomass = c('8:0', '10:0', '10:0 2OH', '11:0',
                                         '12:0', '12:0 2OH', '12:0 3OH',
                                         '13:0', '14:0', '14:0 2OH',
                                         '14:0 3OH', '14:1', '15:0',
                                         '15:0 anteiso', '15:0 iso',
                                         '16:0 10me', '16:0 2OH', '16:0 iso',
                                         '16:1 w5c', '16:1 w7c', '16:1 w9c',
                                         '17:0', '17:0 iso', '17:0 anteiso',
                                         '17:0 cyclo', '17:1', '17:1 iso',
                                         '18:0', '18:0 10me', '18:1 w9c',
                                         '18:1 w9t', '18:2 w6,9c', '19:0',
                                         '19:0 cyclo', '19:1', '20:0')
)

indicator_list
```
